Mixed effect tweedie model in TMB

### 2019-10-21 analysis/tmb-sensor-explore/2-fit-temperature.Rmd
refine plot-map functions: remove utm axis lables from faceted maps and make bathymetry lines finer
update climate models/predictions

d_trawl <- d_trawl[!is.na(d_trawl$depth), ]
d_trawl <- d_trawl[!is.na(d_trawl$temperature_c), ]
d_trawl <- d_trawl[!is.na(d_trawl$SST), ]
d_trawl <- d_trawl[!is.na(d_trawl$meanSST), ]
d_trawl <- mutate(d_trawl,
  DOY = lubridate::yday(date),
  DOY_scaled = arm::rescale(lubridate::yday(date)),
  DOY_scaled2 = arm::rescale(lubridate::yday(date))^2,
  #depth_scaled = arm::rescale(log(depth_m)),
  SST_scaled = arm::rescale(meanSST),
  #depth_scaled2 = depth_scaled^2,
  depth_scaled3 = depth_scaled^3,
  shallow = ifelse(depth>35,0,1),
    deep = ifelse(depth<35,0,1)
)
# d_trawl[is.na(d_trawl$shallow), ]
# d_trawl[is.na(d_trawl$SST_scaled), ]

@@

d_trawl <- d_trawl %>% dplyr::filter(year>2003)
```

@@

```{r}
d_trawl1 <- d_trawl %>% filter(ssid == 3)
spde <- make_spde(d_trawl1$X, d_trawl1$Y, n_knots = 150)
plot_spde(spde)
sum(d_trawl1$shallow)
```

```{r}
d_trawl1$year_scaled <- arm::rescale(d_trawl1$year)
d_trawl1$x_scaled <- arm::rescale(d_trawl1$X)
d_trawl1$y_scaled <- arm::rescale(d_trawl1$Y)

m_temp_SST <- sdmTMB(d_trawl1,
  temperature_c ~ year_scaled + poly(SST_scaled,3)*depth_scaled , #+ SST_scaled*x_scaled + SST_scaled*y_scaled
  #time_varying = ~ 0 + depth_scaled2,
  ar1_fields = FALSE, include_spatial = FALSE,
  time = "year", spde = spde, family = gaussian(link = "identity"),
  silent = FALSE
)

p_SST <- predict(m_temp_SST)

p_SST$residuals <- residuals(m_temp_SST)

ggplot(p_SST, aes(depth_scaled, residuals)) +
  geom_point(alpha=0.4) +
  ylim(-8,8) +
  geom_smooth()

@@

m <- gbm( temperature_c ~ depth_scaled + X + Y, data = d_trawl1, n.trees = 2000, interaction.depth = 3, shrinkage = 0.02)

m2 <- gbm( temperature_c ~ SST_scaled + depth_scaled + X + Y, data = d_trawl1, n.trees = 2000, interaction.depth = 3, shrinkage = 0.02)

@@

```{r}

p_SST <- predict(m_temp_SST)
p <- predict(m_temp)
#predictions3b <- predict(m_temp3_nosst)

p_SST$residuals <- residuals(m_temp_SST)
p$residuals <- residuals(m_temp)

# nd <- data.frame(sst_resid = predictions3$data$residuals, nosst_resid = predictions3b$data$residuals, shallow = d_trawl$shallow, depth = d_trawl$depth)
# 
# ggplot(filter(nd, shallow == 1), aes(nosst_resid,sst_resid, colour = as.factor(shallow))) + 
#   geom_point() + geom_abline(intercept = 0, slope = 1)
```
Residual plots:
```{r}
ggplot(p_SST, aes(est, residuals)) +
  geom_point() +
  geom_smooth() +
  ylim(-8,8) +
  xlim(2,15) +
  facet_wrap(~year)
ggplot(p, aes(est, residuals)) +
  geom_point() +
  geom_smooth() +
  ylim(-8,8) +
   xlim(2,15) +
  facet_wrap(~year)
```

```{r}
ggplot(p_SST, aes(depth_scaled, residuals)) +
  geom_point() +
  geom_smooth() +
  ylim(-7,8) +
  facet_wrap(~year)
ggplot(p, aes(depth_scaled, residuals)) +
  geom_point() +
  geom_smooth() +
  ylim(-7,8) +
  facet_wrap(~year)
```

```{r}
p_SST$depth_mean <- exp(p_SST$depth_scaled*0.66+5)
glimpse(d_trawl)
ggplot(d_trawl, aes(depth_scaled, r)) +
  geom_point(alpha=0.4) +
  ylim(-8,8) +
  geom_smooth()
```

```{r}
p1 <- ggplot(p_SST, aes(depth_scaled, residuals)) +
  geom_point() +
  ylim(-8,8) +
  geom_smooth()
p2 <- ggplot(p, aes(depth_scaled, residuals)) +
  geom_point() +
  ylim(-8,8) +
  geom_smooth()
gridExtra::grid.arrange(p1, p2, nrow = 2)
```

@@

Current top model
```{r}
plan(multiprocess)
temp_rw <- sdmTMB_cv(
  data = d_trawl, n_knots = 130, x = "X", y = "Y", k_folds = 10,
  formula = temperature_c ~ 0 + as.factor(year) + SST_scaled,
  time_varying = ~ 0 + depth_scaled + depth_scaled2,
  time = "year", family = gaussian(link = "identity"),
  ar1_fields = FALSE, include_spatial = TRUE, seed = 999
)
sum(temp_rw$sum_loglik)
```

```{r}
spde <- make_spde(d_trawl$X, d_trawl$Y, n_knots = 220)
plot_spde(spde)
```

```{r}
m_temp_rw <- sdmTMB(d_trawl,
  temperature_c ~ 0 + as.factor(year),
  time_varying = ~ 0 + depth_scaled + depth_scaled2,
  ar1_fields = FALSE, include_spatial = TRUE,
  time = "year", spde = spde, family = gaussian(link = "identity"),
  silent = TRUE
)

predictions_rw <- predict(m_temp_rw)
predictions_rw$residuals <- residuals(m_temp_rw)
cor(predictions_rw$est, predictions_rw$temperature_c)^2 #
# aic(m_temp3_rw)
```

@@

```{r}
m_temp3 <- sdmTMB(d_trawl,
  temperature_c ~ 0 + as.factor(year) + SST_scaled*depth_scaled + SST_scaled*depth_scaled2 + SST_scaled*depth_scaled3,
  ar1_fields = FALSE, include_spatial = TRUE,
  time = "year", spde = spde, family = gaussian(link = "identity"),
  silent = TRUE
)

@@

```{r}
ggplot(predictions3, aes(est, residuals)) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~year)
```

```{r}
ggplot(predictions3, aes(depth_scaled, residuals)) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~year)
```

```{r}
ggplot(predictions3, aes(depth_scaled, residuals)) +
  geom_point() +
  geom_smooth()
```

@@

```{r}
p_nd <- readRDS("../VOCC/data/predicted_temp_allyears.rds")
# View(p_nd)
```

@@

omega_s_all <- plot_map_raster(p_nd, "omega_s") +
  ggtitle("Spatial random effects only") +
   scale_fill_gradient2( low = "Steel Blue 4",
 high = "Red 3", limits = c(-1.5,2.5))+
  labs(fill = "") + 
gfplot::theme_pbs() + theme(axis.title.x = element_blank(), axis.title.y = element_blank(), legend.position = c(.82,.75))

@@

```{r}
 plot_map_raster(p_nd, "epsilon_st") +
  ggtitle("Spatiotemporal random effects only") +
  facet_wrap(~year) +
  scale_fill_gradient2( low = "Steel Blue 4",
 high = "Red 3")
```

```{r}
p_nd1 <- p_nd %>% filter(ssid!=16) %>% filter(ssid!=4)
st_1n3 <- plot_map_raster(p_nd1, "epsilon_st") +
  ggtitle("Spatiotemporal random effects") +
  facet_wrap(~year) +
  scale_fill_gradient2(
    low = "Steel Blue 4",
    high = "Red 3", 
    limits = c(-1,1), 
    na.value = "red") +
  labs(fill = "epsilon") + 
  gfplot::theme_pbs() + 
  theme(
  legend.position = "none", 
  axis.title.x = element_blank(), axis.title.y = element_blank())
st_1n3
```

```{r}
p_nd4 <- p_nd %>% filter(ssid==4) #%>% filter(ssid!=4)
st_4 <- plot_map_raster(p_nd4, "epsilon_st") +
  #ggtitle(" ") +
  facet_wrap(~year) +
  scale_fill_gradient2( low = "Steel Blue 4",
 high = "Red 3", limits = c(-1,1)) +
          labs(fill = " ") + 
gfplot::theme_pbs() + theme(axis.title.x = element_blank(), axis.title.y = element_blank(), legend.position = c(.85,.15))
st_4
```

```{r}
p_nd16 <- p_nd %>% filter(ssid==16) #%>% filter(ssid!=4)
st_16 <- plot_map_raster(p_nd16, "epsilon_st") +
  #ggtitle(" ") +
  facet_wrap(~year) +
  scale_fill_gradient2( low = "Steel Blue 4",
 high = "Red 3", limits = c(-1,1)) +
  xlim(65,345) +
    ylim(5800,6100) +
gfplot::theme_pbs() + theme(axis.title.x = element_blank(), axis.title.y = element_blank(), legend.position = "none")
st_16
```

>>>>

d_trawl <- d_trawl[!is.na(d_trawl$depth), ]
d_trawl <- d_trawl[!is.na(d_trawl$temperature_c), ]
#d_trawl <- d_trawl[!is.na(d_trawl$SST), ]
#d_trawl <- d_trawl[!is.na(d_trawl$meanSST), ]
d_trawl <- mutate(d_trawl,
  DOY = lubridate::yday(date),
  DOY_scaled = arm::rescale(lubridate::yday(date)),
  DOY_scaled2 = arm::rescale(lubridate::yday(date))^2,
  #depth_scaled = arm::rescale(log(depth_m)),
  SST_scaled = arm::rescale(meanSST),
  #depth_scaled2 = depth_scaled^2,
  depth_scaled3 = depth_scaled^3,
  shallow = ifelse(depth>35,0,1),
  deep = ifelse(depth<35,0,1)
)
# d_trawl[is.na(d_trawl$shallow), ]
# d_trawl[is.na(d_trawl$SST_scaled), ]

@@

#d_trawl <- d_trawl %>% dplyr::filter(year>2003)
```

@@

@@

m <- gbm( temperature_c ~ depth_scaled + X + Y, data = d_trawl, n.trees = 2000, interaction.depth = 3, shrinkage = 0.02)
m2 <- gbm( temperature_c ~ SST_scaled + depth_scaled + X + Y, data = d_trawl, n.trees = 2000, interaction.depth = 3, shrinkage = 0.02)

@@

@@

Current top model

```{r}
spde <- make_spde(d_trawl$X, d_trawl$Y, n_knots = 400)
plot_spde(spde)
```

```{r}
m_temp_rw <- sdmTMB(d_trawl,
  temperature_c ~ 0 + as.factor(year),
  time_varying = ~ 0 + depth_scaled + depth_scaled2,
  ar1_fields = TRUE, include_spatial = TRUE, # changed AR1_field to true
  time = "year", spde = spde, family = gaussian(link = "identity"),
  silent = FALSE
)

saveRDS(m_temp_rw, file = (here::here("analysis/tmb-sensor-explore/models/model-temp-all-years-400kn.rds")))

```{r}
predictions_rw <- predict(m_temp_rw)
predictions_rw$residuals <- residuals(m_temp_rw)
cor(predictions_rw$est, predictions_rw$temperature_c)^2 #
# aic(m_temp3_rw)

@@

```{r eval=FALSE}
m_temp3 <- sdmTMB(d_trawl,
  temperature_c ~ 0 + as.factor(year) + SST_scaled*depth_scaled + SST_scaled*depth_scaled2 ,
  ar1_fields = FALSE, include_spatial = TRUE,
  time = "year", spde = spde, family = gaussian(link = "identity"),
  silent = TRUE
)

@@

```{r eval=FALSE}
ggplot(predictions3, aes(est, residuals)) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~year)
```

```{r eval=FALSE}
ggplot(predictions3, aes(depth_scaled, residuals)) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~year)
```

```{r eval=FALSE}
ggplot(predictions3, aes(depth_scaled, residuals)) +
  geom_point() +
  geom_smooth()
```

@@

```{r}
nd_all <- readRDS(paste0("../VOCC/data/nd_all_synoptic.rds")) %>% filter(year<2019)
unique(nd_all$year)
glimpse(nd_all)
predict_temp <- predict(m_temp_rw, newdata = nd_all)
saveRDS(predict_temp, file = "../VOCC/data/predicted_temp_allyears_revised")
p_nd <- predict_temp
#p_nd <- readRDS("../VOCC/data/predicted_temp_allyears.rds")
# View(p_nd)
```

@@

omega_s_all <- plot_map_raster(p_nd, "omega_s") +
  ggtitle("Spatial random effects only") +
    scale_fill_gradient2( low = "Steel Blue 4",
 high = "Red 3", limits = c(-1.5,1.5),
    na.value = "red") +  labs(fill = "") + 
gfplot::theme_pbs() + theme(axis.title.x = element_blank(), axis.title.y = element_blank(), legend.position = c(.82,.75))

@@

```{r}
 plot_map_raster(p_nd, "epsilon_st") +
  ggtitle("Spatiotemporal random effects only") +
  facet_wrap(~year) +   scale_fill_gradient2( low = "Steel Blue 4",
 high = "Red 3", limits = c(-1.5,1.5),
    na.value = "red")
```

```{r}
p_nd1 <- p_nd %>% filter(ssid!=16) %>% filter(ssid!=4)
st_1n3 <- plot_map_raster(p_nd1, "epsilon_st") +
  ggtitle("Spatiotemporal random effects") +
  facet_wrap(~year) +
   scale_fill_gradient2( low = "Steel Blue 4",
 high = "Red 3", limits = c(-1.5,1.5),
    na.value = "red") +
  labs(fill = "epsilon") + 
  gfplot::theme_pbs() + 
  theme(
  legend.position = "none", 
  axis.title.x = element_blank(), axis.title.y = element_blank())
st_1n3
```

```{r}
p_nd4 <- p_nd %>% filter(ssid==4) #%>% filter(ssid!=4)
st_4 <- plot_map_raster(p_nd4, "epsilon_st") +
  #ggtitle(" ") +
  facet_wrap(~year) +
     scale_fill_gradient2( low = "Steel Blue 4",
 high = "Red 3", limits = c(-1.5,1.5),
    na.value = "red") +
          labs(fill = " ") + 
gfplot::theme_pbs() + theme(axis.title.x = element_blank(), axis.title.y = element_blank(), legend.position = c(.85,.15))
st_4
```

```{r}
p_nd16 <- p_nd %>% filter(ssid==16) #%>% filter(ssid!=4)
st_16 <- plot_map_raster(p_nd16, "epsilon_st") +
  #ggtitle(" ") +
  facet_wrap(~year) +
   scale_fill_gradient2( low = "Steel Blue 4",
 high = "Red 3", limits = c(-1.5,1.5),
    na.value = "red") +
  xlim(65,345) +
    ylim(5800,6100) +
gfplot::theme_pbs() + theme(axis.title.x = element_blank(), axis.title.y = element_blank(), legend.position = "none")
st_16
```

### 2019-10-21 analysis/tmb-sensor-explore/3-fit-do.Rmd
update climate models/predictions

d_trawl1 <- d_trawl #%>% filter(ssid != 16) %>% filter(ssid != 4)    
spde <- make_spde(d_trawl1$X, d_trawl1$Y, n_knots = 800)
plot_spde(spde)
```

@@

saveRDS(do_model_all_DOY, file = (here::here("analysis/tmb-sensor-explore/models/model-do-all-DOY2.rds")))
time1
do_model_all_DOY
```

```{r}
do_model_DOY <- readRDS(here::here("analysis/tmb-sensor-explore/models/model-do-all-DOY.rds"))
do_model_DOY

get_diag(do_model_DOY, response = "log_do", variable = "depth_scaled")

@@

Predict DO for whole grid using quadratic of predicted temperature values and mean DOY
```{r}
p_nd <- predict(do_model_DOY, newdata = newdata)
p_nd$do_est <- exp(p_nd$est)

@@

@@

```{r}
p_nd <- readRDS((here::here("analysis/tmb-sensor-explore/models/predicted-DO.rds")))

@@

```{r}
p_nd4 <- p_nd %>% filter(ssid==4)
pred_4 <- plot_map_raster(p_nd4, "do_est") +
  scale_fill_viridis_c(labels = labels, breaks = c(0,2,4,6,8), limits = c(0, 9), na.value = "red")+
  facet_wrap(~year) +
  labs(fill = "ml/L DO") +
  #ggtitle(" ")+ 
  theme(legend.position = "none")
pred_4
```

@@

```{r}
p_nd4 <- p_nd %>% filter(ssid==4) #%>% filter(ssid!=4)
st_4 <- plot_map_raster(p_nd4, "epsilon_st") +
  #ggtitle(" ") +
  facet_wrap(~year) +
  scale_fill_gradient2( high = "Steel Blue 4",
  low = "Red 3", limits = c(-0.5,0.5)) +
          labs(fill = "epsilon") + 
gfplot::theme_pbs() + theme(axis.title.x = element_blank(), axis.title.y = element_blank(), legend.position = "none")
st_4
```

@@

```{r}
png(
  file = "figs/do-predictions.png", 
  res = 600,
  units = "in",
  width = 11, 
  height = 6.5
)

>>>>

d_trawl1 <- d_trawl %>% filter(!(year == 2016 & ssid ==4)) #%>% filter(ssid != 16) %>% filter(ssid != 4)    
#test <- d_trawl1 %>% filter(year == 2016 & ssid ==16) 
spde <- make_spde(d_trawl1$X, d_trawl1$Y, n_knots = 800)
plot_spde(spde)
```

@@

saveRDS(do_model_all_DOY, file = (here::here("analysis/tmb-sensor-explore/models/model-do-without-wcvi2016.rds")))
time1
do_model_all_DOY
```

```{r eval=FALSE}
# do_model_DOY <- readRDS(here::here("analysis/tmb-sensor-explore/models/model-do-all-DOY.rds"))
# do_model_DOY

get_diag(do_model_all_DOY, response = "log_do", variable = "depth_scaled")

@@

Predict DO for whole grid using quadratic of predicted temperature values and mean DOY
```{r eval=FALSE}
p_nd <- predict(do_model_DOY, newdata = newdata)
p_nd$do_est <- exp(p_nd$est)

@@

Save new predictions without QCS gap and WCVI 2016 DO
```{r eval=FALSE}
#new temp and do prediction grid for vectors

model1 <- readRDS("~/github/dfo/gfranges/analysis/tmb-sensor-explore/models/model-temp-2008-on-used.rds")
#model2 <- readRDS("~/github/dfo/gfranges/analysis/tmb-sensor-explore/models/model-do-all-DOY-used.rds")
#update model to exclude WCVI 2016 outliers
model2 <- readRDS("~/github/dfo/gfranges/analysis/tmb-sensor-explore/models/model-do-without-wcvi2016.rds")

nd_all <- readRDS(paste0("data/nd_all_synoptic.rds"))
nd <- filter(nd_all, year >2007) %>% filter(year<2019)
pred_temp <- predict(model1, newdata = nd)
predtemp <- pred_temp %>% mutate(temp_scaled = est, temp_scaled2 = est^2, depth_scaled3 = depth_scaled^3) %>% select(-est, -est_non_rf, -est_rf, -omega_s, -zeta_s, -epsilon_st)
predtemp$DOY_scaled <- 0
pred_do <- predict(model2, newdata = predtemp)
pred_do$temp <- pred_do$temp_scaled*1.296 + 6.49

pred_do$log_do <- pred_do$est
pred_do$do_est <- exp(pred_do$log_do)

# may want to filter estimates for depths less than 15 m
# pred_do <- pred_do %>% filter(depth>15)

saveRDS(pred_do, file="~/github/dfo/gfranges/analysis/VOCC/data/predicted-DO-new.rds")

@@

```{r}
#p_nd <- readRDS((here::here("analysis/tmb-sensor-explore/models/predicted-DO.rds")))
p_nd <- readRDS((here::here("analysis/VOCC/data/predicted-DO-new.rds")))

@@

```{r}
p_nd4 <- p_nd %>% filter(ssid==4) %>% filter(year!=2016)
pred_4 <- plot_map_raster(p_nd4, "do_est") +
  scale_fill_viridis_c(labels = labels, breaks = c(0,2,4,6,8), limits = c(0, 9), na.value = "red")+
  facet_wrap(~year) +
  labs(fill = "ml/L DO") +
  #ggtitle(" ")+ 
  theme(legend.position = "none")
pred_4
```

@@

```{r}
p_nd4 <- p_nd %>% filter(ssid==4) %>% filter(year!=2016)
st_4 <- plot_map_raster(p_nd4, "epsilon_st") +
  #ggtitle(" ") +
  facet_wrap(~year) +
  scale_fill_gradient2( high = "Steel Blue 4",
  low = "Red 3", limits = c(-0.5,0.5)) +
          labs(fill = "epsilon") + 
gfplot::theme_pbs() + theme(axis.title.x = element_blank(), axis.title.y = element_blank(), legend.position = "none")
st_4
```

@@

```{r}
png(
  file = "figs/do-predictions-without-wcvi2016.png", 
  res = 600,
  units = "in",
  width = 11, 
  height = 6.5
)
